---
title: "Data Analysis"
date: "12/2/2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{R Libraries}
library(ggplot2) # Data visualization
library(readr) # CSV file I/O, e.g. the read_csv function
library(vcd)
library(lattice)
library(dplyr)
library(lubridate)
library(plyr)
library(tidyverse)
library(caret)
library(readxl)
```

```{r code}
data<- read_csv("loan_orig.csv") # Raw data
```

```{r convert to data frame}
df = data.frame(data)

```

```{r}
#data transformation for some of the attributes of character to numeric
char_to_convert <- c("annual_inc_joint", "mths_since_last_major_derog", "open_acc_6m",
    "open_il_6m", "open_il_12m", "open_il_24m", "mths_since_rcnt_il",
    "total_bal_il", "il_util", "open_rv_12m", "open_rv_24m",
    "max_bal_bc", "all_util", "total_rev_hi_lim", "total_cu_tl",
    "inq_last_12m", "dti_joint", "inq_fi", "tot_cur_bal", "tot_coll_amt")

df <-
  df %>%
  mutate_at(.vars = char_to_convert,.funs = funs(as.numeric) )

```

```{r}
# check the date attributes

date_attributes <- c( "issue_d", "last_pymnt_d", "last_credit_pull_d",
    "next_pymnt_d", "earliest_cr_line", "next_pymnt_d")
df %>%
  select_(.dots = date_attributes) %>% str()

```

```{r}
#Converting the date attribute from character to date format by assigning the 1st of the mont

date_formater <- function(x){
  as.Date(paste0("01-", x), format = "%d-%b-%Y")
  } 

df <-
  df %>%
  mutate_at(.funs = funs(date_formater), .vars = date_attributes)
```

```{r}
#applying 0 to some of the attributes which have NA's as they carry valuable information and will be helpul in modelling  

na_to_zero <-
  c("mths_since_last_delinq", "mths_since_last_record",
    "mths_since_last_major_derog")

df <- 
  df %>%
  mutate_at(.vars = na_to_zero, .funs = funs(replace(., is.na(.), 0)))
```

Some of the potential varibles which are useful in defining the default payments are,
a) loan_status : This attribute defines the status of the loan
b) delinq_2yrs: The number of 30+ days where an individual has past incidence of the deliquency in 2 years
c) mnths_since_last_delinq : The number of months since the individual last deliquency 

```{r}
default_vars <- c("loan_status", "delinq_2yrs", "mths_since_last_delinq")
purrr::map(.x = df[, default_vars], .f = base::unique)
```

The loan status has all the current status of the loans, doing a count of unique loan status

```{r}
data.frame(table(df$loan_status))
```

The Values of the loan_status can be found from the following link https://triamus.github.io/project/lending-club-loan-data-in-r/#correlation

We can define a loan as default from the following loan_status values,

* Default
* Does not meet the credit policy. Status:Charged Off
* In Grace Period
* Late (16-30 days)
* Late (31-120 days)
So we can list of attributes to define the default attribute for loan dataframe,

```{r}
defaulted <- 
  c("Default", 
    "Status:Charged Off", 
    "In Grace Period", 
    "Late (16-30 days)", 
    "Late (31-120 days)")
head(defaulted)
```
Assign boolean indicator for the default attribute

```{r}
df <-
  df %>%
  mutate(default = ifelse(!(loan_status %in% defaulted), FALSE, TRUE))

```

```{r}
table(df$default) / nrow(df)
```

Let us find attributes which have na's in the loan dataFrame

```{r}

na_count <-sapply(df, function(y) sum(length(which(is.na(y)))))
na_count <- data.frame(na_count)
unique_count <- sapply(df, function(x) sum(length(unique(x))))
unique_count <- data.frame(unique_count)
#View(na_count)
#View(unique_count)

```

Based on na and unique count we can remove the following attributes as they are unfit for model fitting , sub-grade attribute is dropped as grade attribute gives us more insights and loan_status has been used to define the default atribute.

```{r}
remove_attributes = c("annual_inc_joint", "dti_joint", "policy_code", "id", "member_id",
    "emp_title", "url", "desc", "title", "open_acc_6m", "open_il_6m", 
    "open_il_12m", "open_il_24m", "mths_since_rcnt_il", "total_bal_il", 
    "il_util", "open_rv_12m", "open_rv_24m", "max_bal_bc", "all_util",
    "total_rev_hi_lim", "inq_fi", "total_cu_tl", "inq_last_12m",
    "verification_status_joint", "next_pymnt_d","sub_grade", "loan_status")
df <- df %>% dplyr::select(-one_of(remove_attributes))

head(df)
  
```

```{r}
give_count <- 
  stat_summary(fun.data = function(x) return(c(y = median(x)*1.06, label = length(x))),
               geom = "text")

# see http://stackoverflow.com/questions/19876505/boxplot-show-the-value-of-mean
give_mean <- 
  stat_summary(fun.y = mean, colour = "darkgreen", geom = "point", 
               shape = 18, size = 3, show.legend = FALSE)
```

```{r}
ggplot(df, aes(grade, loan_amnt)) +
  geom_boxplot(fill = "white", colour = "darkblue", 
               outlier.colour = "red", outlier.shape = 1) +
  give_count +
  give_mean +
  
  facet_wrap(~ default) +
  labs(title="Loan Amount by Grade", x = "Grade", y = "Loan Amount \n")

```

```{r}
ggplot(df, aes(grade, int_rate)) +
  geom_boxplot(fill = "white", colour = "darkblue", 
               outlier.colour = "red", outlier.shape = 1) +
  give_count +
  give_mean +
  
  labs(title="Interest Rate by Grade", x = "Grade", y = "Interest Rate \n") +
  facet_wrap(~ term)

ggplot(df, aes(home_ownership, int_rate)) +
  geom_boxplot(fill = "white", colour = "darkblue", 
               outlier.colour = "red", outlier.shape = 1) +
  give_count +
  give_mean +
  scale_y_continuous(labels = comma) +
  facet_wrap(~ default) +
  labs(title="Interest Rate by Home Ownership", x = "Home Ownership", y = "Interest Rate \n")
```